The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting

### 2025-05-26 manuscript/05-conclusion.qmd
revision submitted

Our findings highlight the importance of collecting and incorporating domain knowledge when building forecasting models. We evaluated both point and probabilistic forecasts using a range of models, from simple univariate approaches like ARIMA to complex models such as LSTM. Notably, we demonstrated that employing complex models does not necessarily lead to more accurate forecasts in the pharmaceutical supply context. Additionally, our research revealed that newly developed foundational time series forecasting models have potential applicability, particularly in environments where advanced analytical skills and knowledge are limited, as often seen in pharmaceutical supply chains in developing countries.

@@

- The use of distribution-based representation of error metrics across different rolling origins or forecast horizons, such as boxplots, to showcase forecast accuracy provides essential insights into variability in accuracy at different point of time in data. Building on this, future work could aim at designing methodologies that identify diagnostic patterns within the distribution of errors. These approaches could suggest model improvements by analyzing error distributions and help practitioners to refine models based on error behavior.

- While current study focus on monthly aggregate consumption at the organizational level, there is potential in exploring more granular forecasts. Future research could investigate daily or weekly forecasting and examine the impacts at hierarchical levels, such as specific sites or health facilities. Investigating how fine-grained temporal and hierarchical forecasts influence inventory management and operational efficiency could yield valuable insights for reducing stockouts, minimizing costs and reducing waste.

>>>>

Our findings highlight the importance of collecting and incorporating domain knowledge when building forecasting models. We evaluated both point and probabilistic forecasts using a range of models, from simple univariate approaches like ARIMA to complex models such as LSTM. We demonstrated that while the use of complex models may improve forecast accuracy in the pharmaceutical supply context, this comes at the cost of significantly increased computational time—an important consideration for resource-constrained settings in low- and middle-income countries (LMICs). Additionally, our research indicates that newly developed foundational time series forecasting models hold promise, particularly in settings with limited computational infrastructure and constrained access to advanced analytical expertise, as often seen in pharmaceutical supply chains in developing countries.

@@

- \textcolor{blue}{Disruptions such as the COVID-19 pandemic and armed conflicts may alter consumption patterns in time series data, introducing irregularities that challenge the assumptions of traditional forecasting methods. In such contexts, it becomes important not only to model observed consumption but also to account for censored demand—i.e., the unmet need that would have been fulfilled under normal conditions and is essential for effective replenishment planning. Future research could focus on developing and evaluating forecasting strategies that are specifically designed to handle disrupted time series and estimate censored demand. In particular, methods that enhance the robustness of probabilistic forecasts in the presence of structural breaks or exogenous shocks are especially relevant. One promising direction is also the development of causal forecasting models that incorporate policy changes, epidemic dynamics, and socio-economic factors—an approach that may be particularly well suited to the realities of pharmaceutical supply chains in low- and middle-income countries (LMICs).}

- \textcolor{blue}{Future research should explore the integration of probabilistic forecasting into inventory management policies, with a focus on evaluating the operational impact of forecast uncertainty. In addition, there is substantial potential in leveraging more granular data—such as daily or weekly consumption patterns—and examining forecasts at hierarchical levels, including individual sites or health facilities. Investigating how fine-grained temporal and spatial forecasts influence decision-making in inventory control could provide valuable insights into reducing stockouts, minimizing holding costs, and decreasing pharmaceutical waste.}

### 2025-05-26 manuscript/main.qmd
revision submitted

---
title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
# author:
#   - name: Arebu Issa Bilal
#     email: arebu.issa@aau.edu.et
#     affiliations:
#         - id: Addis 
#           name: Addis Ababa University
#           department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
#           country: Ethiopia.
#     attributes:
#         corresponding: true
#   - name: Bahman Rostami-Tabar
#     email: rostami-tabarb@cardiff.ac.uk
#     affiliations:
#         - id: carbs
#           name: Cardiff University
#           department: Data Lab for Social Good Reserach Group, Cardiff Business School 
#           country: UK.
#   - name: Harsha Halgamuwe Hewage
#     email: HalgamuweHewageHR@cardiff.ac.uk
#     affiliations:
#         - ref: carbs
#   - name: Umit Sezer Bititci 
#     email: U.S.Bititci@hw.ac.uk
#     affiliations:
#         - id: herio 
#           name: Heriot-Watt University
#           department: Edinburgh Business School, Heriot-Watt University
#           country: UK.
#   - name: Teferi Gedif Fenta
#     email: tgedif@gmail.com
#     affiliations:
#         - ref: Addis

@@

keywords:
  - Forecasting
  - Pharmaceutical supply chain
  - Domain knowlege
  - Forecast accuracy
  - Developing countries
bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
  pdf-engine: pdflatex
  prefer-html: true
  cache: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
    journal:
      name: International Journal of Production Research
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

```{r results, child="04-results.qmd"}

@@

We use R and Python for model development and accuracy evaluation. The paper is written in Quarto, and all code, data, and files will be shared in a GitHub repository upon acceptance for publication.

>>>>

---
title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
author:
  - name: Arebu Issa Bilal
    email: arebu.issa@aau.edu.et
    affiliations:
        - id: Addis
          name: Addis Ababa University
          department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
          country: Ethiopia.
    attributes:
        corresponding: true
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: Cardiff University
          department: Data Lab for Social Good Reserach Group, Cardiff Business School
          country: UK.
  - name: Harsha Halgamuwe Hewage
    email: HalgamuweHewageHR@cardiff.ac.uk
    affiliations:
        - ref: carbs
  - name: Umit Sezer Bititci
    email: U.S.Bititci@hw.ac.uk
    affiliations:
        - id: herio
          name: Heriot-Watt University
          department: Edinburgh Business School, Heriot-Watt University
          country: UK.
  - name: Teferi Gedif Fenta
    email: tgedif@gmail.com
    affiliations:
        - ref: Addis

@@

keywords:
  - Forecasting
  - Pharmaceutical supply chain
  - Domain knowlege
  - Forecast accuracy
  - Developing countries
bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
  pdf-engine: pdflatex
  prefer-html: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo,xcolor}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
        \usepackage{tikz}
        \usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit}
        \tikzset{
          process/.style={
          rectangle,
          draw=black,
          thick,
          text width=4.5cm,
          align=center,
          minimum height=1.4cm
          },
        groupbox/.style={
          draw=black,
          thick,
          dashed,
          inner sep=0.3cm,
          rounded corners,
          label={[black]90:#1}
        },
        arrow/.style={
          thick,
          ->,
          >=Stealth
        }
                }

    journal:
      name: International Journal of Production Research
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

```{r results, child="04-results.qmd"}
```

@@

To enhance transparency and reproducibility, we not only provide data and the code, but also the entire paper that is written in R & Python. All materials to reproduce this paper is available at [Insert Github link once paper accepted](Insert Github link once paper accepted).

The repository contains the raw data, all R and Python scripts used in experiments, the results used in the paper, as well as the quarto files for producing this paper.
